Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-11-04 RMS/submission.tex
drawed part of counterexample figure

%% bare_jrnl_compsoc.tex
%% V1.4a
%% 2014/09/17
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8a or later) with an IEEE
%% Computer Society journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_conf_compsoc.tex,
%%                    bare_jrnl_compsoc.tex, bare_jrnl_transmag.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices and paper sizes can       ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

@@

\usepackage{alltt}
\usepackage{epic}
\usepackage{pstricks}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

@@

\author{Jiaxiang~Liu% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem J. Liu is with the School of 
Software, Tsinghua University, China and LIX, \'Ecole Polytechnique, France\protect\\
E-mail: jiaxiang.liu@hotmail.com}% <-this % stops an unwanted space
}
%\thanks{Manuscript received April 19, 2005; revised September 17, 2014.}}

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic operating system.
Interrupts would be triggered by the clock every $T$, which we call
\emph{interrupt cycle}. When an interrupt request occurs, if the
system is interruptable, i.e. the interrupt mask is cleared, the
handler function $schedule()$ will be invoked; otherwise $schedule()$
will be pending until the interrupt mask becomes cleared.  The
implementation is shown as $schedule()$ in Algorithm~\ref{a:schedule},
where $taskList$ is the list of periodic tasks to be scheduled. We
assume that the list is descending order of priority, and both
variables $taskList$ and $timer$ are global. In this implementation,
there is only one kind of interrupt, the period $T_i$ of each task is
a multiple of $T$, and the tasks are independent, meeting
assumption~(A2).

@@

In Algorithm~\ref{a:schedule}, the handler function $scheduler()$
first updates status of all tasks in $taskList$ via function
$updateStatus()$. Then it traverses the list to execute the ready
tasks one by one, or to do a return when encountering an
interrupted\footnote{Note that the status \textit{INTERRUPT} indicates
  the task is interrupted for the moment, or was interrupted before
  but its execution is not complete yet.} task. The function
$updateStatus()$ updates each task in two steps: firstly, if the task
is running, it becomes interrupted; secondly for the task at its
initiation time, if its previous job is complete, it would be set to
be ready, otherwise it misses its deadline, producing an error. Notice
that $schedule()$ is invoked only when the periodic interrupt is
handled. Its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

In the rest of this paper, we use \emph{scheduling} to refer to the
stage, from the moment when a pending interrupt request is detected,
to the moment when the first should-be-run periodic task starts
executing, i.e. Line~\ref{l:function} in
Algorithm~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ finding and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}), and (iii) the time for switching context
from $schedule()$ to that first should-be-run task. \emph{Switching}
refers to the stage, from the moment when a periodic task completes
its execution, to the moment when the next should-be-run periodic task
starts executing. \emph{Switching time} thus also consists of three
parts: (i) the time for switching context from the finished task back
to $schedule()$, (ii) the time spent by $schedule()$ finding and
setting the \emph{next} should-be-run periodic task, and (iii) the
time for switching context from $schedule()$ to that next
should-be-run task.

@@

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upperbound while increasing (see Line~\ref{l:timer} in
Algorithm~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upperbound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Algorithm~\ref{a:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined:
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Algorithm~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by the recursive
function as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) 
         :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and function \verb|update_with_| updating the status of individual
task (Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Algorithm~\ref{a:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status 
                            : INTERRUPT >
       else < O : PTask |> fi [owise] .
\end{verbatim}
Given a task, if \verb|TIMER|($timer$) can be divided by its period,
this task should be initiated.  In the case where the task should be
initiated, it is set to be \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

@@

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Algorithm~\ref{a:schedule}. It performs updating the status of
$taskList$ and increasing $timer$ by $1$. This part is modeled by
function \verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, finding the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Algorithm~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

@@

\subsection{Scenarios}
\label{ss:results}
We use the following setting for our verification, which is from the 
statistics provided by our industrial partner:
\begin{itemize}
\item The interrupt cycle is $5ms$.
\item The scheduling time ranges from $5{\mu}s$ to $9{\mu}s$, and the
switching time ranges from $2{\mu}s$ to $4{\mu}s$.
\item The initial state is with empty stack, empty \verb|pc|, cleared 
\verb|mask| and cleared \verb|ir|.
\end{itemize}

@@

@@

@@

\hide{
% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{John Doe}
Biography text here.
\end{IEEEbiographynophoto}

>>>>

@@

\usepackage{alltt}
\usepackage{epic}
\usepackage{pstricks}
\usepackage{tikz}
\newcommand{\hide}[1]{\ignorespaces}
\newcommand{\jx}[1]{{\bf Jiaxiang: }#1{ \bf End}}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

@@

\author{Jiaxiang~Liu,
        Min~Zhou,
        Xiaoyu~Song,
        Ming~Gu,
        and~Jiaguang~Sun% <-this % stops a space
\IEEEcompsocitemizethanks{%
\IEEEcompsocthanksitem J. Liu is with the School of Software, Tsinghua
University, Beijing 100084, China, and also with LIX, \'Ecole
Polytechnique, Palaiseau 91120, France.\protect\\ E-mail:
jiaxiang.liu@hotmail.com
\IEEEcompsocthanksitem M. Zhou, M. Gu and J. Sun are with the School
of Software, Tsinghua University, Beijing 100084, China.
\IEEEcompsocthanksitem X. Song is with the Department of Electrical \&
Computer Engineering, Portland State University, Portland, OR
97207-0751, USA.}% <-this % stops an unwanted space
}
%\thanks{Manuscript received April 19, 2005; revised September 17, 2014.}}

@@

\section{The Implementation of RMS}
\label{s:imp}
The implementation is in a realistic avionic operating system.
Interrupts would be triggered by the clock every $T$, which we call
\emph{interrupt cycle}. When an interrupt request occurs, if the
system is interruptable, i.e. the interrupt mask is cleared, the
handler function $schedule()$ will be invoked; otherwise $schedule()$
will be pending until the interrupt mask becomes cleared.  The
implementation is shown as $schedule()$ in Figure~\ref{a:schedule},
where $taskList$ is the list of periodic tasks to be scheduled. We
assume that the list is descending order of priority, and both
variables $taskList$ and $timer$ are global. In this implementation,
there is only one kind of interrupt, the period $T_i$ of each task is
a multiple of $T$, and the tasks are independent, meeting
assumption~(A2).

@@

In Figure~\ref{a:schedule}, the handler function $scheduler()$
first updates status of all tasks in $taskList$ via function
$updateStatus()$. Then it traverses the list to execute the ready
tasks one by one, or to do a return when encountering an
interrupted\footnote{Note that the status \textit{INTERRUPT} indicates
  the task is interrupted for the moment, or was interrupted before
  but its execution is not complete yet.} task. The function
$updateStatus()$ updates each task in two steps: firstly, if the task
is running, it becomes interrupted; secondly for the task at its
initiation time, if its previous job is complete, it would be set to
be ready, otherwise it misses its deadline, producing an error. Notice
that $schedule()$ is invoked only when the periodic interrupt is
handled. Its execution cannot be interrupted when it is updating
status of tasks or finding the next task to execute, however, it can
be interrupted while executing some task (Line~\ref{l:function}). This
allows the execution of $schedule()$ to be nested.

In the rest of this paper, we use \emph{scheduling} to refer to the
stage, from the moment when a pending interrupt request is detected,
to the moment when the first should-be-run periodic task starts
executing, i.e. Line~\ref{l:function} in
Figure~\ref{a:schedule}. Therefore, \emph{scheduling time} mainly
consists of three parts: (i) the time for switching context from the
running task, possibly none, to $schedule()$ when an interrupt request
is handled, (ii) the time spent by $schedule()$ finding and setting
the \emph{first} should-be-run periodic task
(Lines~\ref{l:1stline}-\ref{l:endrun1st} in
Figure~\ref{a:schedule}), and (iii) the time for switching context
from $schedule()$ to that first should-be-run task. \emph{Switching}
refers to the stage, from the moment when a periodic task completes
its execution, to the moment when the next should-be-run periodic task
starts executing. \emph{Switching time} thus also consists of three
parts: (i) the time for switching context from the finished task back
to $schedule()$, (ii) the time spent by $schedule()$ finding and
setting the \emph{next} should-be-run periodic task, and (iii) the
time for switching context from $schedule()$ to that next
should-be-run task.

@@

\hide{
At last, to make our model checkable by untimed model checking, it is
reasonable to reset the global variable $timer$ when it reaches an
upperbound while increasing (see Line~\ref{l:timer} in
Figure~\ref{a:schedule}) in the model. Then $timer$ is of sort
\verb|Timer| and the upperbound would be the least common multiple of
the periods of all tasks.}

The global variable $timer$ is reset when it reaches an upper bound
while increasing, which is not shown in detail in
Figure~\ref{a:schedule} but is reasonable. The upper bound is the
least common multiple of the periods of all tasks. A sort \verb|Timer|
is defined:
\begin{verbatim}
  op [_/_] : Nat NzNat -> Timer [ctor] .
\end{verbatim}

@@

\subsection{Task Initiation}
\label{ss:init}
Periodic tasks are initiated sequentially by the function
$updateStatus()$ in Figure~\ref{a:schedule}, which is treated as an
instantaneous action in our model. It is modeled by the recursive
function as below:
\begin{verbatim}
  op updateStatus_with_ : TaskList Timer 
                            -> TaskList . 
  eq updateStatus null with TIMER = null .
  eq updateStatus (TASK :: L) with TIMER
       = (update TASK with TIMER) 
         :: (updateStatus L with TIMER) .
\end{verbatim}
with \verb|TIMER| the current value of the global variable $timer$,
and function \verb|update_with_| updating the status of individual
task (Lines~\ref{l:startupdate}-\ref{l:endupdate} in
Figure~\ref{a:schedule}):
\begin{verbatim}
  op update_with_ : Object Timer ~> Object .
  ceq update < O : PTask | period : T, 
                           status : ST > 
        with TIMER
      = if ST == DORMANT 
        then < O : PTask | status : READY >
        else error fi
      if TIMER rem T == 0 .
  eq update < O : PTask | status : ST > 
       with TIMER
     = if ST == RUNNING 
       then < O : PTask | status 
                            : INTERRUPT >
       else < O : PTask |> fi [owise] .
\end{verbatim}
Given a task, if \verb|TIMER|($timer$) can be divided by its period,
this task should be initiated.  In the case where the task should be
initiated, it is set to be \verb|READY| if its \verb|status| is
\verb|DORMANT|; otherwise that means the previous job of this task is
not complete, hence it misses its deadline, producing an
\verb|error|. In the other case where the task should not be
initiated, its \verb|status| changes only if it is \verb|RUNNING|.

@@

Unlike periodic tasks, even though the scheduling stage is modeled by
a \verb|Counter|, its functionality is too important to be abandoned.
We divide the behaviors of scheduling into three parts.  The first
part contains its timed behaviors. This part is modeled by regarding
scheduling as a system task of sort \verb|SysTask|. Modeling timed
behaviors of tasks is explained in Section~\ref{ss:timedbehavior}.
The other two parts together define its functionality. The second part
corresponds to Lines~\ref{l:updatestatus}-\ref{l:inc} in
Figure~\ref{a:schedule}. It performs updating the status of
$taskList$ and increasing $timer$ by $1$. This part is modeled by
function \verb|_.startScheduling| which applies instantaneously at the
beginning of scheduling, as shown in rule \verb|interrupt-handle|:
\begin{verbatim}
  op _.startScheduling : System -> System .
  eq (L T STS HW ISRC).startScheduling 
     = ((updateStatus L with T) 
        inc(T) STS HW ISRC) .
\end{verbatim}
The third part corresponds to
Lines~\ref{l:startrun1st}-\ref{l:endrun1st}, finding the first
should-be-run periodic task and setting it to execute. It is modeled
by function \verb|_.finishScheduling|, which applies instantaneously
at the end of scheduling:
\begin{verbatim}
  op _.finishScheduling : System -> System .
  eq (L T STS HW ISRC).finishScheduling
     = (L T (finish scheduling in STS) 
        HW ISRC).run1stTask .
\end{verbatim}
where \verb|finish_in_| resets the counter of task \verb|scheduling|,
and \verb|_.run1stTask| models
Lines~\ref{l:startrun1st}-\ref{l:endrun1st} in
Figure~\ref{a:schedule}, searching the task with highest priority
that has status \verb|INTERRUPT| or \verb|READY| then performing an
\emph{interrupt return} or executing it, respectively.

@@

\subsection{Scenarios}
\label{ss:results}
We use the following setting for our verification, which is from the 
statistics provided by our industrial partner:
\begin{itemize}
\item The interrupt cycle is $5ms$.
\item The scheduling time ranges from $15{\mu}s$ to $38{\mu}s$, and the
switching time ranges from $10{\mu}s$ to $20{\mu}s$.
\hide{\item The scheduling time ranges from $5{\mu}s$ to $9{\mu}s$, and the
switching time ranges from $2{\mu}s$ to $4{\mu}s$.}
\item The initial state is with empty stack, empty \verb|pc|, cleared 
\verb|mask| and cleared \verb|ir|.
\end{itemize}

@@

\begin{figure*}[!t]
\centering
\begin{tikzpicture}[
    % styles
    system task/.style={fill=gray!70,draw=gray},
    periodic task/.style={fill=gray!70!black,draw=gray!50!black},
    task line/.style={gray!30,very thin},
    time line/.style={gray!50,very thin,dashed}]

  \foreach \x in {0.5,1.2,...,3.3}
    \draw[task line] (0,\x) -- (15.7,\x);
  \draw[->,thick] (0,0) -- (16,0);
  \draw[thin] (0,0) -- (0,3.8);

  \draw[time line] (0.5,3.3) -- (0.5,0);
  \draw[time line] (2.5,3.3) -- (2.5,0);
  \filldraw[system task] (0,1.2) rectangle +(0.5,0.2);
  \filldraw[periodic task] (0.5,3.3) rectangle +(2,0.2);
  \filldraw[system task] (2.5,0.5) rectangle +(0.25,0.2);
  \filldraw[periodic task] (2.75,2.6) rectangle +(1.25,0.2);
  \filldraw[system task] (4,0.5) rectangle +(0.25,0.2);
  \filldraw[periodic task] (4.25,1.9) rectangle +(0.75,0.2);

  \filldraw[system task] (5,1.2) rectangle +(0.5,0.2);
  \filldraw[periodic task] (5.5,3.3) rectangle +(2,0.2);
  \filldraw[system task] (7.5,0.5) rectangle +(0.25,0.2);
  \filldraw[periodic task] (7.75,1.9) rectangle +(2.25,0.2);

  \filldraw[system task] (10,1.2) rectangle +(0.5,0.2);
  \filldraw[periodic task] (10.5,3.3) rectangle +(2,0.2);
  \filldraw[system task] (12.5,0.5) rectangle +(0.25,0.2);
  \filldraw[periodic task] (12.75,2.6) rectangle +(1.25,0.2);
  \filldraw[system task] (14,0.5) rectangle +(0.25,0.2);
  \filldraw[periodic task] (14.25,1.9) rectangle +(0.75,0.2);
\end{tikzpicture}
\caption{An Counterexample of Scenario (iv).}
\label{f:tikz}
\end{figure*}

@@

\hide{

@@

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{John Doe}
Biography text here.
\end{IEEEbiographynophoto}
